---
title: "Creating plots with openAlexR"
author: ["Alex Jack"]
date: 2025-11-25
categories: ["R"]
tags: ["Data Management"]
---
  
Writing a thesis or paper introduction? Don't just *say* your field is growing or that prediction is becoming important—**show it**. With a quick Web of Science export and a few lines of R, you can create compelling visualizations that demonstrate trends in your field and precisely why what you're talking about is *important*.

Here's how to go from search results to publication trend plots in about 5 minutes.

Here are the packages you will need to actually run this.

```{r}
library(dplyr) # for general data-wrangling purposes
library(openalexR)
library(ggplot2)
```

Above the `dplyr` package is great for data wrangling and clean 

```{r,cache=TRUE}
results <- oa_fetch(
  entity = "works",
  title.search = "prediction OR predict OR forecasting",  # Search in title
  abstract.search = "prediction OR predict OR decision support tool OR forecasting",  # Search in abstract
  concepts.id = "https://openalex.org/C86803240",  # Ecology concept ID
  from_publication_date = "2015-01-01",
  to_publication_date = "2025-12-31",
  options = list(sample = 100, seed = 12345) # we're going to only sample those that  
)
```

```{r}
# OpenAlex returns a lot of metadata - let's select key columns
df_clean <- results %>%
  select(
    id,
    title = display_name,
    publication_year,
    publication_date,
    type,
    cited_by_count,
    doi,
    language # This is a list column
  ) %>%
  filter(!is.na(publication_year)) %>%
  filter(publication_year >= 2000 & publication_year <= 2025)
```
```{r}
df_yearly <- df_clean %>%
  group_by(publication_year) %>%
  summarise(
    num_publications = n(),
    mean_citations = mean(cited_by_count, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(publication_year)
```
```{r}
ggplot(df_yearly, aes(x = publication_year, y = num_publications)) +
  geom_point(color = "steelblue", size = 3) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_smooth(method = "lm", se = TRUE, 
              linetype = "dashed", color = "coral", linewidth = 0.8) +
  labs(
    x = "Publication Year", 
    y = "Number of Publications",
    title = "Growth of Predictive Ecology Research (2000-2025)",
    subtitle = "Data from OpenAlex: prediction + forecasting + decision support",
    caption = paste("Total unique records:", nrow(df_clean))
  ) +
  theme_minimal(base_size = 12) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5))
```
```{r}
ggplot(df_yearly, aes(x = publication_year, y = mean_citations)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_line(color = "darkgreen", linewidth = 1) +
  labs(
    x = "Publication Year", 
    y = "Mean Citations per Paper",
    title = "Citation Impact by Publication Year",
    subtitle = "More recent papers have fewer citations (less time to accumulate)"
  ) +
  theme_minimal(base_size = 12) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(2015, 2025, by = 1))
```