---
title: "Meta-Analysis on Randomized Clinical Trials - Part 3"
author: ["Amir Golzan"]
date: 2025-06-17 09:00:00 -0800
categories: ["Meta_Analysis", "R"]
tags: ["meta_analysis", "r"] # tags always lowercase
params:
  math: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this part, we begin performing meta-analysis using the R language. If you are not familiar with R, you may refer to [R for Data Science](https://r4ds.had.co.nz/) for a comprehensive introduction.

We will cover how to prepare data for import into RStudio, import the data, label it appropriately, and analyze it using meta-analysis codes.

## Data Preparation

To perform meta-analysis, we first need to calculate the **mean change** and the **Standard Deviation (SD)** of the mean change for studies reporting only pre- and post-measurements. Additionally, we should label our data based on defined criteria (discussed further below) to prepare for subgroup analysis, which is critical in meta-analyses.

Let's say we have extracted the following data from a study evaluating C-reactive protein (CRP) levels in blood after consuming Vitamin E:

```{r table1, echo=FALSE, results='asis'}
library(flextable)

table1 <- data.frame(
  Pre_mean_CRP = "10.7",
  Pre_SD_CRP = "7.9",
  Post_mean_CRP = "8.7",
  Post_SD_CRP = "8.4",
  mean_change_CRP = "?",
  SD_change_CRP = "?"
)

flextable(table1)
```

Now we can calculate the mean change and SD using the following formulas:

**Mean Change** = Post Mean - Pre Mean

**SD of Change**= \(\sqrt{(SD_1^2 + SD_2^2) - (2 \times r \times SD_1 \times SD_2)}\)

For this formula, *r* is calculated using following formula in which Pre_SD and Post_SD should be taken from a known study in which authors themselves reported SD Change.

**r** = \( \frac{\text{Pre}_\text{SD}^2 + \text{Post}_\text{SD}^2 - \text{SD}_\text{change}^2}{2 \times \text{Pre}_\text{SD} \times \text{Post}_\text{SD}} \)

In our example, we assume *r* as 0.98. Considering these, we can now calculate `mean_change_CRP` and `SD_change_CRP`:

```{r table2, echo=FALSE, results='asis'}
library(flextable)

table2 <- data.frame(
  Pre_mean_CRP = "10.7",
  Pre_SD_CRP = "7.9",
  Post_mean_CRP = "8.7",
  Post_SD_CRP = "8.4",
  mean_change_CRP = "-2",
  SD_change_CRP = "1.70"
)

flextable(table2)
```

## Subgroup Analysis Preparation

To perform subgroup analysis, we must assign codes to variables based on which we want to conduct the analysis. Common variables include:

- **Gender**
- **Dosage**
- **Duration of intervention**
- **Health status of individuals**
- **Adjustment for baseline values**
- **Study design**
- **Adherence to the intervention**

For each variable, create a column in Excel and assign codes. For example, for gender: Both genders = Code 1, Male = Code 2, Female = Code 3. These codes facilitate subgroup analysis to understand how variables like gender affect outcomes.

## Meta-Analysis

The first analysis we will become familiar with is the main meta-analysis, which we can perform in R using `metacont` from the `meta` package. We highly recommend running `help(meta)` as it provides comprehensive information on how to run your analyses and which codes to use.

Below, we demonstrate a meta-analysis using a sample dataset. In this sample dataset, we want to understand how using Vitamin E affects the levels of CRP in blood. You can download the data to your local computer from [Sample Data](https://training.cochrane.org/handbook).

```{r meta_analysis, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
# ── 0. Packages ───────────────────────────────────────────────
library(readr)
library(meta)

# ── 1.  Read the data ─────────────────────────────────────────
sample_data <- read_csv("sample_data.csv", show_col_types = FALSE)   

# ── 2.  Main meta-analysis
main_analyis <- metacont(
  ## Variables for Experimental Group  (n_int, mean_change_CRP_int, SD_change_CRP_int)
  n.e      = n_int,                 ## ⇨ “Number of participants for Exp. Group: n”
  mean.e   = mean_change_CRP_int,   ## ⇨ “mean”
  sd.e     = SD_change_CRP_int,     ## ⇨ “sd”

  ## Vars for Control Group  (n_con, mean_change_CRP_con, SD_change_CRP_con)
  n.c      = n_con,                 ## ⇨ “Number of participants for Control Group: n”
  mean.c   = mean_change_CRP_con,   ## ⇨ “mean”
  sd.c     = SD_change_CRP_con,     ## ⇨ “sd”

  ## Labels  (Name = authr name)
  studlab  = `author name`,         ## ⇨ “Labels for Data  ▸ Name”
  data     = sample_data,           ## ⇨ “Selecting data”

  ## Statistic  (No Standard → Weighted Mean Difference)
  ## If studies used different units, use SMD.
  sm       = "MD",                ## ⇨ ‘Tells R no standardization needed—units are the same.’

  ## Pooling model  (Fixed, Inverse-Variance)
  ## We'll discuss this later—using Fixed model for now.
  comb.fixed  = TRUE,               ## ⇨ ‘Pooling Model: Fixed, Inverse-Variance’
  comb.random = FALSE,              ## (leave random OFF)
)

# ── 3.  Review numeric results ─────────────────────
summary(main_analyis)     # pooled WMD, I², τ², Q-test. 

```

In this sample data, the effect of Vitamin E on blood CRP levels in adults were analyzed (this is demonstration-only data). The `common effect model` represents the result of pooling all studies together. The pooled estimate is `-0.40 [-0.52; -0.27]`, which means that consuming Vitamin E may reduce CRP levels in the blood by approximately `0.40 mg/L`. The next question is: **Is this finding statistically significant?** The numbers within brackets indicate the 95% confidence interval (CI). If this interval includes **zero**, the finding is not statistically significant. If it does not include zero, the finding is significant at the 95% confidence level (which we specified for this analysis). In our example, the confidence interval does not include zero, indicating that the result is statistically significant. The p-value further confirms this, being less than 0.0001. 

Heterongenity is also a very important concept in meta-anlysis studies. Heterogeneity refers to differences among studies. When heterogeneity is high or statistically significant, it means that the included studies differ considerably in certain characteristics (like methodology).

> **Note**: The less variation there is between the studies (i.e., the lower the heterogeneity), the more valid and precise the findings from the meta-analysis will be.

So how do we assess heterogeneity? We usually use the following criteria as a general rule of thumb:

- ***Based on I-squared (I^2):***
  - **Less than 50: Low heterogeneity**
  - **Greater than 50: High heterogeneity**
  - **Value of zero: No heterogeneity**

- ***Based on P-value for heterogeneity (from Cochran’s Q test):***
  - **P-value less than 0.05 or 0.1 indicates significant heterogeneity**

In our example, the I^2 is 66.8% and the p-value is 0.0003, indicating significant heterogeneity across studies. We can use I^2 as an indicator to help decide whether to use a fixed or random effects model.

- **Fixed Model:** This model assumes that the studies are similar to each other or that there is low heterogeneity among them. In other words, selecting this model in the analysis causes the R to perform the meta-analysis under the assumption of low heterogeneity. Therefore, this model should be used when heterogeneity is low (I^2 below 50%).

- **Random Model::** This model assumes that the studies are different from each other or that there is high heterogeneity among them. In other words, selecting this model in the analysis causes the R to perform the meta-analysis under the assumption of high heterogeneity. Therefore, this model should be used when heterogeneity is high (I^2 above 50%). 

In our example, since we have high heterogeneity, we should use a random effects model.

```{r meta_analysis_2, echo=TRUE, results='asis', message=FALSE, warning=FALSE}

main_analyis_2 <- metacont(
  ## Variables for Experimental Group
  n.e      = n_int,
  mean.e   = mean_change_CRP_int,
  sd.e     = SD_change_CRP_int,

  ## Vars for Control Group
  n.c      = n_con,
  mean.c   = mean_change_CRP_con,
  sd.c     = SD_change_CRP_con,

  ## Labels  (Name = authr name)
  studlab  = `author name`,
  data     = sample_data,

  ## Statistic  (No Standard → Weighted Mean Difference)
  sm       = "MD",

  ## Pooling model
  comb.fixed  = FALSE,
  comb.random = TRUE,
)

# ── 3.  Review numeric results ─────────────────────
summary(main_analyis_2)

```

The results changed under the random model and are no longer statistically significant. In other words, Vitamin E consumption leads to a non-significant reduction in CRP levels by 0.43 mg/L. This analysis was conducted under the assumption that the studies are heterogeneous. Note that the level of heterogeneity remains similar in both the Fixed and Random models.

## Visualization

The main way to visualize a meta-analysis is by using a forest plot. We will use the `forest()` function from the `meta` package, as shown in the code below:

```{r visualization, echo=TRUE, results='asis', message=FALSE, warning=FALSE}
# ── 4.  Forest plot ──────
forest(main_analyis_2,
       # ── layout options ────────────────────────────────────
       leftcols   = c("studlab"),                      # nly the author column
       rightcols  = c("effect", "ci", "w.fixed"),      # WMD, 95 % CI, weight
       rightlabs  = c("WMD",   "95% CI", "Weight"),

       # ── cosmetic tweaks (optional) ────────────────────────
       sortvar    = year,        # same chronological order as before
       xlab       = "Weighted Mean Difference (Δ CRP, mg/L)",
       fs.study   = 8,           # shrink fonts if you need more rows
       fs.axis    = 8)

```

You can modify your plot using the various options offered by the `forest()` function. To see all available options, run `?forest` in your console.

---

This seems like a lot for now! In the next part, we will explore the sources of heterogeneity, which will lead us to performing subgroup analyses.

**Stay tuned!**

---